---
title: "Alpha Diversities"
output:
  html_document:
    toc: true
    toc_float: true
---

Below are descriptions of common diversity metrics and brief explanations on
their use, along with some test calculations

Minimal calculations of these metrics will be used with the `esophagus` data
found in `phyloseq`.

```{r}
library(phyloseq)
library(vegan)
data("esophagus")
```

**Table of Contents**

- [Observed](#observed)
- [Chao1](#chao1)

Observed
========

This diversity metric simply calculates the number of non-zero abundant taxa.

```{r}
# Use phyloseq function
estimate_richness(esophagus, measures = "Observed")

# Confirm with manual calculation
esoTF <- otu_table(esophagus)
colSums(esoTF != 0) # Non-zero taxa will be TRUE
```

Chao1
=====

According to original paper, the Chao1 metric is a

> Nonparametric estimation of the number of classes in a population

This metric is defined as

$$S_{est} = S_{obs} + \frac{f_1^2}{2f_2}$$

where

- $S_{est}$ is the estimated number of species,
- $S_{obs}$ is the observed number of species (i.e. number of non-zero taxa),
- $f_1$ is the number of singletons (i.e. number of taxa occurring only once),
  and
- $f_2$ is the number of doubletons (i.e. number of taxa occurring only twice).

The abstract of the original paper mentions that this metric estimates number of
classes

> ... when most of the information is concentrated on the low order occupancy
> numbers. (Chao, 1984)

So this implies you should use this metric if you believe there are low abundant
taxa involved in your samples.

The potential rationale for weighting more with singletons and doubletons is
because if you have a lot of singletons, it may suggest that there are more
unobserved taxa in your population. Thus, having a higher diversity.

**Note**: The `vegan` implementation of the calculation adds in correction
terms (the `+1` for the doubletons) in case there are no doubletons in your
sample. Other wise, you'd be dividing by zero. 

```{r}
# Use phyloseq function
estimate_richness(esophagus, measures = "Chao1")
estimateR(t(otu_table(esophagus)), "chao")["S.chao1", ] # Function used in vegan

# Confirm with manual calculation
chao1 <- function(otu) {
    temptable <- otu_table(otu)
    singletons <- colSums(temptable == 1) # Number of singletons
    doubletons <- colSums(temptable == 2) # Number of doubletons
    obs <- colSums(temptable != 0) # Observed number of non-zero taxa

    # Chao1 calculation
    obs + (singletons * (singletons - 1)) / (2 * (doubletons + 1))
}
chao1(otu_table(esophagus))
```

**References**

Chao, Anne. "Nonparametric estimation of the number of classes in a population." Scandinavian Journal of statistics (1984): 265-270. URL: http://www.jstor.org/stable/4615964

Chao1 calculation in the `estimateR()` function within `vegan`
URL: https://github.com/vegandevs/vegan/blob/master/R/estimateR.default.R

Alpha Diversity http://www.coloss.org/beebook/I/gut-symbionts/2/2/4
